Automatic Identification and Annotation of Tasks in Structured Programs

### 2018-04-03 docs/ACM/paper.tex
Done with Section 3.

This paper describes \Taskminer, a source-to-source compiler that exposes task
parallelism in C/C++ programs.
To fulfil this goal, \Taskminer{} solves different challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to the memory blocks accessed within
those regions (Section~\ref{sub:sra}).
Third, it extracts parameters from the code to estimate when it is profitable to
create tasks.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:rec}).
Fourth, \Taskminer{} determines which program variables need to be 
privatized in new tasks, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing
readable annotations (Section~\ref{sub:ir}).

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_cutoff}
\caption{Challenge~\ref{ch:cutoff}: bounding the creation of recursive tasks with
counter associated with function calls.}
\label{fig:ex_cutoff}
\end{center}
\end{figure}

@@

We introduce Challenge~\ref{ch:cutoff} by quoting Duran {\em et al.}:
``{\em In task parallel languages, an important factor for achieving a good
performance is the use of a cut-off technique to reduce the number of tasks
created}"~\cite{Duran08b}.
This observation is particularly true in the context of recursive, fine-grain,
tasks, as we analyze in Section~\ref{sub:rec}.
Figure~\ref{fig:ex_cutoff} provides an example.
To place a limit on the number of tasks simultaneously in flight, we associate
the invocation of recursive functions annotated with task pragmas with a
counter -- \textsf{taskminer\_depth\_cutoff} in Figure~\ref{fig:ex_cutoff}.
The guard in line 7 ensures that we never exceed \textsf{DEPTH\_CUTOFF}, a
predetermined threshold.
This example, together with Figure~\ref{fig:ex_Regions}, lets us emphasize that
the code generation algorithms presented in this paper are
parameterized by constants such as \textsf{DEPTH\_CUTOFF}, or
\textsf{WORK\_CUTOFF} in Figure~\ref{fig:ex_Regions}. Although these
have default estimates, we provide them as parameters to be set at will.

@@

The program in Figure~\ref{fig:ex_expansion} (a), which is a slight variation of
the code earlier seen in Figure~\ref{fig:ex_Regions}, contains two windmills.
The first consists in the \textsf{for} loop at line 1.
The second, nested in the first, is the \textsf{for} loop at line 2.
The vanes that sprout from these windmills, are highlighted in the program
dependence graph seen in Figure~\ref{fig:ex_expansion} (b).
The figure shows that we have three task candidates, each one formed by a
different vane.
In order to find the actual tasks, we apply function \textsf{expand}
(Fig.~\ref{fig:expand_alg}) onto the innermost vanes: \textsf{Vane-2} and
\textsf{Vane-3}.
The latter cannot be further expanded, because it depends on the former.
On the other hand, \textsf{Vane-2} can be expanded to include also
\textsf{Vane-3}, as this expansion does not create new dependences between vanes
in the same windmill.

\paragraph{Rational:}
We expand tasks up to a certain threshold \textsf{COST}, which is determined
by the cost model mentioned in Section~\ref{sub:profit}.
If the profitability of a task is less than this constant, then parallelism
might not pay off the cost of managing threads.
However, if we expand tasks too much, then we might not have much parallelism,
because code within a task is executed sequentially.
Thus, determining \textsf{COST} is essential to the performance of our
optimization.
In this work, we have determined this parameter empirically for the
runtime environment to be described in Section~\ref{sec:eval}.

@@

\paragraph{Bounding Recursive Tasks}
Although the OpenMP runtime guarantees correctness, its scheduling policy
is heavily jeopardized by the excessive queueing up of tasks. Specifically, algorithms
such as \emph{SparseLU} and \emph{FFT} can suffer bloat because of
too many tasks in the Task Graph\cite{Huynh17}.

To solve this problem, we conservatively limit the number of tasks
that can be created tied to the same Task Region. We do that by two means:
(1) limiting the creation of recursive tasks by depth; (2) limiting the creation of 
tasks in the whole program. We do that by setting up static variables to count up
the depth level in recursive calls. When the depth reaches a certain level, which
can be defined by the user, task dispatch will be halted, avoiding creation of tasks that are
too small. This technique is illustrated in Figure \ref{fig:ex_cutoff}. 
This parameter's default value is very conservative and so can be easily switched off
or set by the user.

>>>>

This paper describes \Taskminer, a source-to-source compiler that exposes task
parallelism in C/C++ programs.
To fulfil this goal, \Taskminer{} solves different challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to the memory blocks accessed within
those regions (Section~\ref{sub:sra}).
Third, it extracts parameters from the code to estimate when it is profitable to
create tasks.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}) and limit their recursion
depth (Section~\ref{sub:ir}).
Fourth, \Taskminer{} determines which program variables need to be 
privatized in new tasks, or shared among them (Section~\ref{sub:variance}).
Finally, it maps all this information back into source code, producing
readable annotations (Section~\ref{sub:ir}).

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_cutoff}
\caption{Challenge~\ref{ch:cutoff}: bounding the creation of recursive tasks.
Example taken from~\cite[Fig.1]{Iwasaki16}.}
\label{fig:ex_cutoff}
\end{center}
\end{figure}

@@

We introduce Challenge~\ref{ch:cutoff} by quoting Duran {\em et al.}:
``{\em In task parallel languages, an important factor for achieving a good
performance is the use of a cut-off technique to reduce the number of tasks
created}"~\cite{Duran08b}.
This observation is particularly true in the context of recursive, fine-grain,
tasks, as we analyze in Section~\ref{sub:ir}.
Figure~\ref{fig:ex_cutoff} provides an example.
To place a limit on the number of tasks simultaneously in flight, we associate
the invocation of recursive functions annotated with task pragmas with a
counter -- \textsf{taskminer\_depth\_cutoff} in Figure~\ref{fig:ex_cutoff}.
The guard in line 7 ensures that we never exceed \textsf{DEPTH\_CUTOFF}, a
predetermined threshold.
This example, together with Figure~\ref{fig:ex_Regions}, lets us emphasize that
the code generation algorithms presented in this paper are
parameterized by constants such as \textsf{DEPTH\_CUTOFF}, or
\textsf{WORK\_CUTOFF} in Figure~\ref{fig:ex_Regions}. Although these
have default estimates, we provide them as parameters to be set at will.

@@

The program in Figure~\ref{fig:ex_expansion} (a), a slight variation of
the code seen in Figure~\ref{fig:ex_Regions}, contains two windmills.
The first consists of the \textsf{for} loop at line 1.
The second, nested in the first, is the \textsf{for} loop at line 2.
The vanes that sprout from these windmills are highlighted in the program
dependence graph in Figure~\ref{fig:ex_expansion} (b).
The figure shows that we have three task candidates, each formed by a
different vane.
To find the actual tasks, we apply function \textsf{expand}
(Fig.~\ref{fig:expand_alg}) onto the innermost vanes: \textsf{Vane-2} and
\textsf{Vane-3}.
The latter cannot be expanded, as it depends on the former.
On the other hand, \textsf{Vane-2} can be expanded to include
\textsf{Vane-3}, as this expansion does not create new dependences between vanes
in the same windmill.

\noindent
\textbf{Rational:}
We expand tasks up to the \textsf{COST} threshold, which is determined
by the cost model seen in Section~\ref{sub:profit}.
If the profitability of a task is less than this constant, then parallelism
might not pay off the cost of managing threads.
However, if we expand tasks too much, then we might lose parallelism,
as code in a task runs sequentially.
Thus, determining \textsf{COST} is essential to the performance of our
optimization.
In this work, we have set this parameter empirically for the
runtime environment to be described in Section~\ref{sec:eval}.

@@

\paragraph{Bounding Recursive Tasks}
During the evaluation of the \Taskminer, we observed that recursive
programs experienced performance regressions due to the excessive creation
of tasks.
To avoid this kind of slowdown, we currently give users the possibility to
bound the number of threads ever in flight, via a command line option, e.g.,
\textsf{./Taskminer -r 12} will limit the number of tasks to 12.
We implement this feature directly at the source code level, as part of the
final annotation of code.
Our solution is simple, yet, as the reader shall perceive in
Section~\ref{sec:eval}, it brings non-negligible benefits onto recursive
benchmarks.
Task bounding is implemented via a global variable, statically linked in the
program.
This variable is \textsf{taskminer\_depth\_cutoff} in Figure~\ref{fig:ex_cutoff}.
We insert code to increment it at the beginning of recursive functions that
are invoked within task regions, and we insert code to decrement it at each
return point of said functions.

\begin{example}[Task Bounding]
Figure~\ref{fig:ex_cutoff} illustrates the strategy that we
use to limit the number of tasks in flight due to recursive function
invocation.
The parameter \textsf{DEPTH\_CUTOFF} is determined by \Taskminer's users.
\end{example}

There are more involved ways to bound the number of tasks.
We believe that the state-of-the-art in the field today is the work of
Iwasaki and Taura~\cite{Iwasaki16,Iwasaki16B}.
These authors propose different techniques to limit the creation of tasks,
be it through the replication of code, be it through the estimation of work,
given function inputs.
Pragmas for the conditional creation of tasks, such as the one seen in
lines 7 and 10 of Figure~\ref{fig:ex_cutoff}, lets us obtain much of the benefit
of code versioning, as proposed by Iwasaki and Taura.
However, we found work estimation of recursive functions a task too difficult
to accomplish in general.
Quoting~\cite[p.355]{Iwasaki16}: ``{\em there are tasks which essentially do not
have simple termination conditions (e.g., tasks traversing pointer-based
trees)}".
Thus, although simple, our recursion counters are general enough, handling even
such tasks that are hard to bound symbolically.
Nevertheless, we still would like to explore further ways to cut-off extremely
fine-grained tasks in the future.
